---
title: "Calibration Management"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibration Management}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

This guide shows you how to save, load, and reuse SEC calibrations with measure.sec. Key tasks covered:

1. Save a calibration from a prepped recipe
2. Load and reuse a saved calibration
3. Inspect calibration quality
4. Compare calibrations over time

## Setup

```{r setup}
library(measure)
library(measure.sec)
library(recipes)
library(dplyr)
library(ggplot2)
```

## How to Create and Save a Calibration

### Step 1: Fit a Calibration

First, create a recipe with calibration standards and prep it to fit the calibration:

```{r create-calibration}
# Load polystyrene standards
data(sec_ps_standards)

# View the standards
sec_ps_standards |>
  select(standard_name, mp, log_mp, retention_time) |>
  head(8)
```

```{r fit-calibration}
# Prepare standards with required column names
# The step expects 'retention' and 'log_mw' columns
ps_cal_standards <- sec_ps_standards |>
  select(retention = retention_time, log_mw = log_mp)

# Create a simple data frame for the recipe
# (In practice, this would be your chromatogram data)
sample_data <- tibble(
  sample_id = "test",
  elution_time = seq(10, 20, by = 0.01),
  ri_signal = dnorm(seq(10, 20, by = 0.01), mean = 14, sd = 1)
)

# Create and prep a recipe with conventional calibration
rec <- recipe(~ ., data = sample_data) |>
  step_measure_input_long(
    ri_signal,
    location = vars(elution_time),
    col_name = "ri"
  ) |>
  step_sec_conventional_cal(
    standards = ps_cal_standards,
    fit_type = "cubic"
  )

# Prep the recipe (this fits the calibration)
prepped <- prep(rec)
```

### Step 2: Save the Calibration

Use `save_sec_calibration()` to extract and save the calibration for reuse:

```{r save-calibration}
# Create a temporary file for this example
cal_file <- tempfile(fileext = ".rds")

# Save with metadata for traceability
save_sec_calibration(
 prepped,
  cal_file,
  metadata = list(
    column = "PLgel 5um Mixed-C",
    instrument = "Agilent 1260",
    analyst = "JW",
    notes = "Monthly calibration - January 2024"
  )
)
```

### Format Options

Two formats are supported:

| Format | Extension | Best For |
|--------|-----------|----------|
| RDS | `.rds` | Default. Preserves all R objects exactly |
| YAML | `.yaml` | Human-readable, good for version control |

```{r save-yaml, eval=FALSE}
# Save as YAML for human readability
save_sec_calibration(prepped, "calibration.yaml")
```

## How to Load and Reuse a Calibration

### Load a Saved Calibration

```{r load-calibration}
# Load the calibration
cal <- load_sec_calibration(cal_file)

# View calibration details
print(cal)
```

### Use in a New Recipe

Pass the loaded calibration to `step_sec_conventional_cal()` to skip fitting:

```{r reuse-calibration}
# New sample data
new_samples <- tibble(
  sample_id = "unknown_polymer",
  elution_time = seq(10, 20, by = 0.01),
  ri_signal = dnorm(seq(10, 20, by = 0.01), mean = 15, sd = 0.8)
)

# Create recipe using saved calibration (no fitting needed)
rec_new <- recipe(~ ., data = new_samples) |>
  step_measure_input_long(
    ri_signal,
    location = vars(elution_time),
    col_name = "ri"
  ) |>
  step_sec_conventional_cal(calibration = cal)

# Prep and bake
prepped_new <- prep(rec_new)
result <- bake(prepped_new, new_data = NULL)
```

## How to Inspect Calibration Quality

### View Fit Diagnostics

```{r inspect-diagnostics}
# Summary shows per-standard results
summary(cal)
```

### Key Quality Metrics

| Metric | Good Value | Meaning |
|--------|------------|---------|
| R² | > 0.999 | Fit explains variance |
| RMSE (log MW) | < 0.05 | Prediction error in log units |
| Max % deviation | < 5% | Worst individual standard |

### Visualize the Calibration Curve

```{r plot-calibration, fig.height=4}
# Plot calibration curve from standards data
plot_sec_calibration(sec_ps_standards)
```

```{r plot-with-residuals, fig.height=6}
# With residuals panel to check for systematic errors
plot_sec_calibration(sec_ps_standards, show_residuals = TRUE)
```

## When to Recalibrate vs. Reuse

### Reuse Existing Calibration When:

- Same column and mobile phase
- Column performance is stable (check plate count)
- Within your lab's validated timeframe (typically 1-4 weeks)
- Running routine samples

### Recalibrate When:

- New column installed
- Mobile phase composition changed
- Column performance has degraded
- Significant time has passed
- Method validation requires fresh calibration
- Results seem inconsistent

## Best Practices

### Naming Conventions

Use descriptive, consistent file names:

```r
# Good: includes key information
"ps_plgel-mixed-c_2024-01-15.rds"
"pmma_superose6_instrument2_2024-01.yaml"

# Avoid: ambiguous names
"calibration.rds"
"cal_new.rds"
```

### Include Metadata

Always add metadata for traceability:

```{r metadata-example, eval=FALSE}
save_sec_calibration(
  prepped,
  "ps_calibration.rds",
  metadata = list(
    column = "PLgel 5um Mixed-C, S/N ABC123",
    instrument = "Agilent 1260 #2",
    analyst = "Jane Doe",
    date = Sys.Date(),
    mobile_phase = "THF + 0.1% BHT",
    flow_rate = "1.0 mL/min",
    temperature = "35C",
    notes = "Monthly recalibration after column flush"
  )
)
```

### Version Control (YAML Format)

For labs using git or similar version control:

```{r yaml-vc, eval=FALSE}
# YAML format is diff-friendly
save_sec_calibration(prepped, "calibrations/ps_current.yaml")

# Git will show meaningful diffs when calibration changes
```

### Lab Workflow Integration

A typical workflow:

1. **Monthly**: Run calibration standards, save new calibration with date
2. **Daily**: Load current calibration for sample analysis
3. **Archive**: Keep old calibrations for historical reference
4. **Compare**: Periodically compare calibrations to detect drift

```{r workflow-example, eval=FALSE}
# Monthly calibration routine
monthly_cal <- prep(calibration_recipe)
save_sec_calibration(
  monthly_cal,
  sprintf("calibrations/ps_%s.rds", format(Sys.Date(), "%Y-%m")),
  metadata = list(analyst = Sys.info()["user"])
)

# Daily sample analysis
cal <- load_sec_calibration("calibrations/ps_2024-01.rds")
samples_recipe <- recipe(...) |>
  step_sec_conventional_cal(calibration = cal) |>
  prep()
```

## Cleanup

```{r cleanup}
# Remove temporary file
unlink(cal_file)
```

## See Also

- [Getting Started](getting-started.html) - Basic SEC workflow and concepts
- [System Suitability](system-suitability.html) - Validate column performance
- [Multi-Detector SEC](triple-detection.html) - Absolute MW without calibration
- [Exporting Results](exporting-results.html) - Summary tables and reports

## Session Info

```{r session-info}
sessionInfo()
```
