---
title: "Copolymer Composition Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Copolymer Composition Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

SEC with dual detection (UV and RI) can reveal compositional heterogeneity in copolymers. By analyzing the UV/RI ratio across the molecular weight distribution, you can:

- Detect compositional drift during polymerization
- Identify blocky vs random copolymer structures
- Quantify the composition at different molecular weights

This vignette covers:

1. UV/RI ratio analysis
2. Composition calculations
3. Interpreting compositional heterogeneity

## Setup

```{r setup}
library(measure)
library(measure.sec)
library(recipes)
library(dplyr)
library(ggplot2)
```

## Copolymer Detection Principles

### UV and RI Response

Different monomers have different detector responses:

| Monomer Type | UV Response | RI Response |
|--------------|-------------|-------------|
| Aromatic (styrene) | Strong | Moderate |
| Aliphatic (acrylate) | Weak/None | Moderate |
| UV-absorbing chromophore | Strong | Variable |
| Non-absorbing | None | Based on dn/dc |

### UV/RI Ratio Interpretation

```{r uv-ri-concept, echo = FALSE, fig.width=8, fig.height=4}
# Simulate different copolymer types
time <- seq(8, 14, by = 0.05)
n_points <- length(time)

# Homogeneous copolymer (flat ratio)
homo_ri <- dnorm(time, mean = 11, sd = 0.8)
homo_uv <- 0.5 * homo_ri
homo_ratio <- rep(0.5, n_points)

# Gradient copolymer (ratio changes across MW)
gradient_ri <- dnorm(time, mean = 11, sd = 0.8)
gradient_uv <- 0.5 * gradient_ri * (1 + 0.3 * (time - 11))
gradient_ratio <- gradient_uv / gradient_ri

copolymer_data <- tibble(
  time = rep(time, 2),
  ratio = c(homo_ratio, gradient_ratio),
  type = rep(c("Homogeneous (random)", "Compositional drift"), each = n_points)
)

ggplot(copolymer_data, aes(time, ratio, color = type)) +
  geom_line(linewidth = 1.2) +
  labs(
    x = "Elution Time (min)",
    y = "UV/RI Ratio",
    title = "UV/RI Ratio Profiles",
    subtitle = "Flat ratio = uniform composition; changing ratio = compositional heterogeneity",
    color = "Copolymer Type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Example Dataset

The package includes `sec_copolymer`, a dataset of styrene-acrylate copolymers with varying compositions:

```{r data}
data(sec_copolymer, package = "measure.sec")

# View the dataset structure
glimpse(sec_copolymer)

# Available samples
sec_copolymer |>
  distinct(sample_id, styrene_fraction, description)
```

## UV/RI Ratio Analysis

### Basic Workflow

```{r uv-ri-ratio}
# Analyze a single copolymer sample
copoly_40 <- sec_copolymer |>
  filter(sample_id == "Copoly-40S")

# UV/RI ratio analysis for compositional heterogeneity
rec_ratio <- recipe(
  ri_signal + uv_254_signal + elution_time ~ sample_id,
  data = copoly_40
) |>
  update_role(sample_id, new_role = "id") |>
  # Convert signals to measure format
  step_measure_input_long(
    ri_signal,
    location = vars(elution_time),
    col_name = "ri"
  ) |>
  step_measure_input_long(
    uv_254_signal,
    location = vars(elution_time),
    col_name = "uv"
  ) |>
  # Baseline correction
  step_sec_baseline(measures = c("ri", "uv")) |>
  # Calculate UV/RI ratio
  step_sec_uv_ri_ratio(
    uv_col = "uv",
    ri_col = "ri",
    smooth = TRUE          # Apply smoothing for noise reduction
  )

prepped_ratio <- prep(rec_ratio)
result_ratio <- bake(prepped_ratio, new_data = NULL)

# View result columns
names(result_ratio)
```

### Extracting the Ratio Curve

```{r extract-ratio}
# The uv_ri_ratio column contains a measure_list
# Extract the ratio values at each elution time
ratio_data <- result_ratio |>
  select(sample_id, uv_ri_ratio) |>
  tidyr::unnest(uv_ri_ratio)

# Plot ratio vs elution time
ggplot(ratio_data, aes(location, value)) +
  geom_line(color = "#A23B72") +
  labs(
    x = "Elution Time (min)",
    y = "UV/RI Ratio",
    title = "UV/RI Ratio Profile for 40% Styrene Copolymer",
    subtitle = "Flat ratio indicates uniform composition across MW"
  ) +
  theme_minimal()
```

## Composition Calculation

### Using Known Response Factors

When you know the response factors for each monomer, calculate actual composition:

```{r composition}
# Composition calculation with known response factors for styrene-acrylate
rec_comp <- recipe(
  ri_signal + uv_254_signal + elution_time ~ sample_id,
  data = copoly_40
) |>
  update_role(sample_id, new_role = "id") |>
  step_measure_input_long(ri_signal, location = vars(elution_time), col_name = "ri") |>
  step_measure_input_long(uv_254_signal, location = vars(elution_time), col_name = "uv") |>
  step_sec_baseline(measures = c("ri", "uv")) |>
  # Calculate composition from UV/RI
  step_sec_composition(
    uv_col = "uv",
    ri_col = "ri",
    # Response factors for styrene (UV-active)
    component_a_uv = 1.0,    # Strong UV absorption at 254 nm
    component_a_ri = 0.185,  # dn/dc in THF
    # Response factors for acrylate (UV-inactive)
    component_b_uv = 0.02,   # Weak UV absorption
    component_b_ri = 0.084   # dn/dc in THF
  )

prepped_comp <- prep(rec_comp)
result_comp <- bake(prepped_comp, new_data = NULL)

# Result contains composition columns
result_comp |>
  select(sample_id, starts_with("composition_"))
```

### Response Factor Determination

Response factors should be measured experimentally:

1. **dn/dc**: Measure with differential refractometer
2. **UV extinction**: Measure with UV-Vis spectrophotometer

Typical values (THF, 25°C):

| Polymer | dn/dc (mL/g) | ε₂₅₄ (mL/(g·cm)) |
|---------|--------------|-------------------|
| Polystyrene | 0.185 | 1.0 |
| PMMA | 0.084 | ~0.01 |
| Polybutadiene | 0.127 | 0.05 |
| PEG | 0.069 | ~0 |

## Interpreting Results

### Uniform Composition

```{r uniform-concept, echo = FALSE, fig.width=7, fig.height=5}
time <- seq(8, 14, by = 0.05)

# Uniform copolymer
ri_signal <- dnorm(time, mean = 11, sd = 0.8)
composition <- rep(0.45, length(time))

plot_data <- tibble(
  time = time,
  ri = ri_signal,
  composition = composition
)

ggplot(plot_data) +
  geom_area(aes(time, ri), alpha = 0.3, fill = "#2E86AB") +
  geom_line(aes(time, ri), color = "#2E86AB", linewidth = 1) +
  geom_line(aes(time, composition), color = "#A23B72", linewidth = 1.2) +
  scale_y_continuous(
    name = "RI Signal",
    sec.axis = sec_axis(~., name = "Weight Fraction A")
  ) +
  labs(
    x = "Elution Time (min)",
    title = "Uniform Random Copolymer",
    subtitle = "Flat composition profile across MW distribution"
  ) +
  theme_minimal()
```

**Characteristics:**
- Constant UV/RI ratio
- Flat composition profile
- Indicates random or alternating copolymer

### Compositional Drift

```{r drift-concept, echo = FALSE, fig.width=7, fig.height=5}
# Gradient copolymer
composition_drift <- 0.35 + 0.2 * (time - 11) / 3

plot_data <- tibble(
  time = time,
  ri = ri_signal,
  composition = pmax(0.2, pmin(0.7, composition_drift))
)

ggplot(plot_data) +
  geom_area(aes(time, ri), alpha = 0.3, fill = "#2E86AB") +
  geom_line(aes(time, ri), color = "#2E86AB", linewidth = 1) +
  geom_line(aes(time, composition), color = "#A23B72", linewidth = 1.2) +
  scale_y_continuous(
    name = "RI Signal",
    sec.axis = sec_axis(~., name = "Weight Fraction A")
  ) +
  labs(
    x = "Elution Time (min)",
    title = "Compositional Drift",
    subtitle = "High MW chains enriched in component A (styrene)"
  ) +
  theme_minimal()
```

**Characteristics:**
- Changing UV/RI ratio
- Sloped composition profile
- Indicates gradient copolymer or reactivity differences

### Bimodal Composition

```{r bimodal-concept, echo = FALSE, fig.width=7, fig.height=5}
# Two polymer populations
ri_bimodal <- 0.6 * dnorm(time, mean = 10, sd = 0.6) +
              0.4 * dnorm(time, mean = 12, sd = 0.5)
uv_bimodal <- 0.8 * dnorm(time, mean = 10, sd = 0.6) +
              0.1 * dnorm(time, mean = 12, sd = 0.5)
composition_bimodal <- uv_bimodal / (ri_bimodal + 0.01)
composition_bimodal <- composition_bimodal / max(composition_bimodal) * 0.6

plot_data <- tibble(
  time = time,
  ri = ri_bimodal,
  composition = composition_bimodal
)

ggplot(plot_data) +
  geom_area(aes(time, ri), alpha = 0.3, fill = "#2E86AB") +
  geom_line(aes(time, ri), color = "#2E86AB", linewidth = 1) +
  geom_line(aes(time, composition), color = "#A23B72", linewidth = 1.2) +
  scale_y_continuous(
    name = "RI Signal",
    sec.axis = sec_axis(~., name = "Weight Fraction A")
  ) +
  labs(
    x = "Elution Time (min)",
    title = "Bimodal Composition",
    subtitle = "Two distinct polymer populations with different compositions"
  ) +
  theme_minimal()
```

**Characteristics:**
- Abrupt ratio changes
- Step-like composition profile
- Indicates blend or block copolymer

## Comparing Multiple Copolymers

Analyze all copolymer samples to compare UV/RI ratios across compositions:

```{r compare-copolymers, fig.width=8, fig.height=5}
# Analyze all samples (excluding homopolymers for this example)
copolymers_only <- sec_copolymer |>
  filter(stringr::str_starts(sample_id, "Copoly"))

# Process all copolymers
rec_all <- recipe(
  ri_signal + uv_254_signal + elution_time ~ sample_id,
  data = copolymers_only
) |>
  update_role(sample_id, new_role = "id") |>
  step_measure_input_long(ri_signal, location = vars(elution_time), col_name = "ri") |>
  step_measure_input_long(uv_254_signal, location = vars(elution_time), col_name = "uv") |>
  step_sec_baseline(measures = c("ri", "uv")) |>
  step_sec_uv_ri_ratio(uv_col = "uv", ri_col = "ri", smooth = TRUE)

prepped_all <- prep(rec_all)
result_all <- bake(prepped_all, new_data = NULL)

# Extract ratio curves for plotting
all_ratios <- result_all |>
  select(sample_id, uv_ri_ratio) |>
  tidyr::unnest(uv_ri_ratio)

# Plot UV/RI ratios - higher ratio = more styrene
ggplot(all_ratios, aes(location, value, color = sample_id)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Elution Time (min)",
    y = "UV/RI Ratio",
    color = "Sample",
    title = "UV/RI Ratio Profiles by Styrene Content",
    subtitle = "Higher ratio corresponds to more styrene (UV-absorbing)"
  ) +
  scale_color_viridis_d() +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Best Practices

### Detector Calibration

1. **Verify response factors** with homopolymer standards
2. **Account for detector delay** between UV and RI
3. **Use same solvent** as for response factor determination

### Data Quality

1. **Adequate signal-to-noise** on both detectors
2. **Baseline correction** before ratio calculation
3. **Signal threshold** to avoid noise in low-signal regions

### Interpretation Caveats

1. **UV/RI ratio is relative**, not absolute composition
2. **Assumes no specific interactions** between detector and polymer
3. **Block copolymers** may show different behavior than random

## See Also

- [Getting Started](getting-started.html) - Basic SEC workflow and concepts
- [Multi-Detector SEC](triple-detection.html) - Detector setup and processing
- [Calibration Management](calibration-management.html) - Save and reuse calibrations
- [Exporting Results](exporting-results.html) - Summary tables and report generation

## Session Info

```{r session-info}
sessionInfo()
```
