---
title: "Exporting Results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exporting Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

After processing SEC data with measure.sec, you'll need to export results for reports, further analysis, or regulatory submissions. This guide covers:

1. Creating summary tables with MW averages
2. Extracting slice-by-slice data for detailed analysis
3. Comparing multiple samples side-by-side
4. Generating automated reports
5. Exporting to Excel, CSV, and other formats

## Quick Reference: Export Functions

| Function | Purpose | Output |
|----------|---------|--------|
| `measure_sec_summary_table()` | Per-sample MW averages | Tibble with Mn, Mw, Mz, dispersity |
| `measure_sec_slice_table()` | Point-by-point data | Long or wide format tibble |
| `measure_sec_compare()` | Multi-sample comparison | Summary, differences, optional plot |
| `measure_sec_report()` | Automated reports | HTML, PDF, or Word document |

## Setup

```{r setup}
library(measure)
library(measure.sec)
library(recipes)
library(dplyr)
library(ggplot2)
```

## Creating Processed Data for Export

In a real workflow, you'd process SEC data through calibration and MW averaging steps (see the [Getting Started](getting-started.html) vignette). Here, we'll create representative processed data to demonstrate the export functions.

```{r create-mock-data}
# Simulate processed SEC results with realistic MW values
# In practice, these columns come from step_sec_mw_averages()

processed <- tibble::tibble(
  sample_id = c("PS-50K", "PS-100K", "PS-200K"),
  polymer_type = "polystyrene",
  # MW averages (typical for narrow PS standards)
  mw_mn = c(48500, 97200, 195000),
  mw_mw = c(50200, 101000, 203000),
  mw_mz = c(52100, 105000, 211000),
  mw_dispersity = c(1.035, 1.039, 1.041),
  # Additional metadata
  injection_volume = 100,  # µL
  concentration = 2.0      # mg/mL
)

processed
```

For multi-sample comparisons, let's create data for different polymer types:

```{r create-comparison-data}
# Three batches with slight variations (typical QC scenario)
batch1 <- tibble::tibble(
  sample_id = "Batch-001",
  mw_mn = 48200, mw_mw = 50100, mw_mz = 52000, mw_dispersity = 1.039
)

batch2 <- tibble::tibble(
  sample_id = "Batch-002",
  mw_mn = 49100, mw_mw = 51200, mw_mz = 53100, mw_dispersity = 1.043
)

batch3 <- tibble::tibble(
  sample_id = "Batch-003",
  mw_mn = 47800, mw_mw = 49800, mw_mz = 51900, mw_dispersity = 1.042
)
```

## Summary Tables

### Basic Summary Table

`measure_sec_summary_table()` creates a one-row-per-sample table with key metrics:

```{r summary-table}
# Create summary table - automatically finds MW columns
summary_tbl <- measure_sec_summary_table(
  processed,
  sample_id = "sample_id"
)

print(summary_tbl)
```

### Including Additional Columns

You can include any numeric column from your data:

```{r summary-with-extra}
# Include method metadata alongside MW results
summary_with_meta <- measure_sec_summary_table(
  processed,
  sample_id = "sample_id",
  additional_cols = c("injection_volume", "concentration")
)

print(summary_with_meta)
```

### Controlling Decimal Places

For regulatory submissions that require specific precision:

```{r summary-digits}
# Higher precision for documentation
summary_precise <- measure_sec_summary_table(
  processed,
  sample_id = "sample_id",
  digits = 0  # Whole numbers for MW
)

print(summary_precise)
```

## Slice Tables: Point-by-Point Data

### Understanding Slice Data

SEC analysis works on a point-by-point basis across the chromatogram. Each "slice" represents one data point with its elution time and corresponding signal values. Use `measure_sec_slice_table()` to extract this detailed data from recipes that produce `measure_list` columns.

### Creating Slice Data

First, let's process actual chromatogram data to demonstrate slice extraction:

```{r process-for-slices}
# Load and process actual chromatogram data
data(sec_triple_detect)

# Process a single sample
ps_sample <- sec_triple_detect |>
  filter(sample_id == "PS-100K")

rec <- recipe(
  ri_signal + elution_time ~ sample_id,
  data = ps_sample
) |>
  update_role(sample_id, new_role = "id") |>
  step_measure_input_long(
    ri_signal,
    location = vars(elution_time),
    col_name = "ri"
  ) |>
  step_sec_baseline(measures = "ri")

chromatogram_data <- prep(rec) |> bake(new_data = NULL)
```

### Long Format (Default)

Long format is best for plotting with ggplot2:

```{r slice-long}
# Extract slice data in long format
slices_long <- measure_sec_slice_table(
  chromatogram_data,
  measures = "ri",
  sample_id = "sample_id"
)

# View structure - one row per data point
head(slices_long, 10)
```

```{r plot-slices, fig.height=4}
# Plot using the slice table
ggplot(slices_long, aes(x = location, y = value)) +
  geom_line(color = "#2E86AB") +
  labs(
    x = "Elution Time (min)",
    y = "RI Signal (baseline corrected)",
    title = "Chromatogram from Slice Data"
  ) +
  theme_minimal()
```

### Wide Format

Wide format is better for spreadsheet export or correlating multiple measures:

```{r slice-wide}
# Wide format puts each measure in its own column
slices_wide <- measure_sec_slice_table(
  chromatogram_data,
  measures = "ri",
  sample_id = "sample_id",
  pivot = TRUE
)

head(slices_wide)
```

### Exporting Slice Data to CSV

```{r export-csv, eval=FALSE}
# Export for external analysis
write.csv(slices_long, "sec_slice_data.csv", row.names = FALSE)

# Or use readr for consistent formatting
readr::write_csv(slices_long, "sec_slice_data.csv")
```

## Comparing Multiple Samples

### Basic Comparison

`measure_sec_compare()` provides side-by-side comparison with differences from a reference:

```{r compare-basic}
# Compare the three batches we created earlier
comparison <- measure_sec_compare(
  batch1, batch2, batch3,
  samples = c("Batch 001", "Batch 002", "Batch 003"),
  metrics = "mw_averages",
  plot = FALSE
)

print(comparison)
```

### Understanding the Comparison Output

The comparison object contains:

- **`$summary`**: All samples with their metrics
- **`$differences`**: Absolute and percent differences from reference
- **`$plot`**: MWD overlay (if requested and available)
- **`$reference`**: Which sample is the reference

```{r compare-details, eval=FALSE}
# Access individual components
comparison$summary      # Metrics for all samples
comparison$differences  # Differences from reference
comparison$reference    # Reference sample name
```

### Setting a Different Reference

By default, the first sample is the reference. You can change this:

```{r compare-reference, eval=FALSE}
# Use PS 500K as reference
comparison_ref <- measure_sec_compare(
  ps100k, ps500k, pmma,
  samples = c("PS 100K", "PS 500K", "PMMA High"),
  reference = "PS 500K"
)
```

### Batch-to-Batch Comparison

A common use case is comparing production batches to a reference lot:

```{r batch-comparison, eval=FALSE}
# Compare production batches
batch_comparison <- measure_sec_compare(
  reference_lot,
  batch_001,
  batch_002,
  batch_003,
  samples = c("Reference", "Batch 001", "Batch 002", "Batch 003"),
  reference = "Reference"
)

# Check for significant deviations
batch_comparison$differences |>
  filter(abs(mw_mw_pct) > 5)  # Flag batches > 5% different
```

## Automated Reports

### Available Templates

```{r list-templates}
# See available report templates
list_sec_templates()
```

### Standard Report

The standard template includes:
- Summary table with MW averages
- Chromatogram overlay
- Molecular weight distribution plot

```{r generate-report, eval=FALSE}
# Generate HTML report
measure_sec_report(
  processed,
  template = "standard",
  output_format = "html",
  title = "Polymer SEC Analysis",
  author = "Lab Analyst"
)
```

### Detailed Report

For more comprehensive documentation:

```{r detailed-report, eval=FALSE}
measure_sec_report(
  processed,
  template = "detailed",
  output_format = "pdf",
  title = "Comprehensive SEC Report",
  sample_id = "sample_id",
  include_slice_table = TRUE  # Append raw data
)
```

### QC Report

For system suitability testing:

```{r qc-report, eval=FALSE}
measure_sec_report(
  sst_data,
  template = "qc",
  output_format = "html",
  specs = list(
    plate_count_min = 10000,
    asymmetry_min = 0.8,
    asymmetry_max = 1.5,
    resolution_min = 1.5
  )
)
```

### Saving Reports to Specific Locations

```{r report-location, eval=FALSE}
# Save to a specific path
measure_sec_report(
  processed,
  template = "standard",
  output_file = "reports/polymer_analysis_2024-01-15.html",
  open = FALSE  # Don't open automatically
)
```

## Integration with Other Tools

### Export to Excel

For spreadsheet-based workflows:

```{r excel-export, eval=FALSE}
# Requires writexl package
library(writexl)

# Create a workbook with multiple sheets
write_xlsx(
  list(
    "Summary" = summary_tbl,
    "Slice Data" = slices_long
  ),
  "sec_results.xlsx"
)
```

### Export to Database

For LIMS integration:

```{r db-export, eval=FALSE}
# Using DBI for database connection
library(DBI)

con <- dbConnect(RSQLite::SQLite(), "lims.db")

# Write summary to database
dbWriteTable(con, "sec_results", summary_tbl, append = TRUE)

dbDisconnect(con)
```

### Export for GraphPad Prism

Prism prefers wide format with specific column arrangements:

```{r prism-export, eval=FALSE}
# Create Prism-friendly format
prism_data <- slices_long |>
  select(location, sample_id, value) |>
  tidyr::pivot_wider(names_from = sample_id, values_from = value)

write.csv(prism_data, "sec_for_prism.csv", row.names = FALSE)
```

### Export for Python Analysis

For interoperability with Python workflows:

```{r python-export, eval=FALSE}
# Save as feather for fast Python loading
arrow::write_feather(slices_long, "sec_data.feather")

# Or as parquet for columnar storage
arrow::write_parquet(slices_long, "sec_data.parquet")
```

## Best Practices

### Traceability

Always include metadata for regulatory compliance:

```{r traceability, eval=FALSE}
# Add audit trail information
summary_with_audit <- summary_tbl |>
  mutate(
    analysis_date = Sys.Date(),
    analyst = "JW",
    instrument = "Agilent 1260",
    column = "PLgel Mixed-C",
    software_version = packageVersion("measure.sec")
  )
```

### File Naming Conventions

Use descriptive, consistent file names:

```r
# Good: includes key information
"sec_summary_batch123_2024-01-15.xlsx"
"mwd_comparison_stability_t12m.pdf"

# Avoid: ambiguous names
"results.xlsx"
"output.csv"
```

### Version Control for Reports

Save report parameters for reproducibility:

```{r version-control, eval=FALSE}
# Save analysis parameters alongside results
analysis_params <- list(
  date = Sys.time(),
  calibration = "ps_2024-01.rds",
  baseline_method = "linear",
  integration_limits = c(8, 18),
  package_version = packageVersion("measure.sec")
)

saveRDS(analysis_params, "sec_analysis_params.rds")
```

## Troubleshooting

### Missing Columns in Summary

If MW columns are missing from your summary table, check that you've run the calibration and MW averaging steps:
```{r troubleshoot-missing, eval=FALSE}
# Ensure you have MW data
names(processed)  # Should include Mn, Mw, Mz, or mw_mn, mw_mw, mw_mz
```

### Empty Slice Tables

If `measure_sec_slice_table()` returns no data:

```{r troubleshoot-empty, eval=FALSE}
# Check that measure columns exist
find_measure_cols <- function(data) {
  names(data)[vapply(data, inherits, logical(1), "measure_list")]
}
find_measure_cols(processed)
```

### Report Generation Fails

If `measure_sec_report()` fails:

1. Verify Quarto is installed: `quarto::quarto_version()`
2. For PDF output, ensure LaTeX is installed
3. Check data has required columns

## Session Info

```{r session-info}
sessionInfo()
```
