---
title: "SEC Concepts: A Practical Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SEC Concepts: A Practical Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## About This Guide

This guide explains SEC concepts from a practical perspective---what you need to know to use `measure.sec` effectively. It's not a textbook; every concept ties back to package functionality with guidance on *when* and *why* to use different approaches.

**Who this is for**: Users with a chemistry background who are new to SEC or want to understand when to use which package features.

**How to use it**: Read straight through for a conceptual foundation, or jump to specific sections when you need decision guidance.

## How SEC Data Flows Through measure.sec

### The Processing Pipeline

SEC data processing follows a consistent pattern: raw detector signals are converted to a common format, processed to remove artifacts, calibrated to molecular weight, and then used to calculate results. Understanding this flow helps you build recipes that make sense.

```{r pipeline-diagram, echo = FALSE, fig.width=8, fig.height=3}
library(ggplot2)

# Create a simplified visual representation of the pipeline using numeric positions
stages <- data.frame(
  x = 1:5,
  y = 1,
  stage = c("Raw Signals", "Preprocessing", "Calibration", "MW Calculation", "Results"),
  description = c(
    "RI, UV, MALS,\nViscometer",
    "Baseline, delay,\nconcentration",
    "Conventional,\nbroad, universal",
    "Mn, Mw, Mz,\ndistributions",
    "Tables, plots,\nreports"
  )
)

ggplot(stages, aes(x = x, y = y)) +
  geom_segment(aes(x = 1.15, xend = 4.85, y = 1, yend = 1),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "#2E86AB", linewidth = 1.5) +
  geom_point(size = 15, color = "#2E86AB") +
  geom_text(aes(label = stage), y = 1.15, size = 3.5, fontface = "bold") +
  geom_text(aes(label = description), y = 0.85, size = 2.8, lineheight = 0.9, color = "gray30") +
  scale_x_continuous(limits = c(0.5, 5.5)) +
  scale_y_continuous(limits = c(0.7, 1.25)) +
  labs(title = "SEC Data Processing Pipeline") +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  )
```

### Stage 1: Input Signals

Every recipe starts by converting raw detector signals into `measure` format using `step_measure_input_long()`. This standardizes different detectors into a common structure where each measurement has a location (elution time or volume) and a value (signal intensity).

```r
recipe(...) |>
  step_measure_input_long(ri_signal, location = vars(elution_time), col_name = "ri") |>
  step_measure_input_long(uv_signal, location = vars(elution_time), col_name = "uv")
```

### Stage 2: Preprocessing

Before calibration, signals need cleaning:

| Step | What It Does | When to Use |
|------|-------------|-------------|
| `step_sec_detector_delay()` | Aligns multi-detector signals | Always with multiple detectors |
| `step_sec_baseline()` | Removes baseline drift | Almost always |
| `step_sec_ri()` / `step_sec_uv()` | Normalizes by response factor | Before concentration calculation |
| `step_sec_concentration()` | Converts to mass concentration | Before light scattering or viscometry |
| `step_sec_band_broadening()` | Corrects peak spreading | High-resolution MW distributions |

**Order matters**: Detector delay correction must come first. Baseline correction should precede normalization. Concentration is needed before MALS calculations.

### Stage 3: Calibration

Calibration converts elution time to molecular weight. The right approach depends on your standards and goals:

| Method | Function | Best For |
|--------|----------|----------|
| Conventional | `step_sec_conventional_cal()` | Narrow standards, same polymer type |
| Broad standard | `step_sec_broad_standard()` | Single characterized reference |
| Universal | `step_sec_universal_cal()` | Different polymer types |
| Absolute (MALS) | `step_sec_mals()` | True MW, no standards needed |

### Stage 4: MW Calculation

With calibration complete, calculate molecular weight results:

| Step | Output |
|------|--------|
| `step_sec_mw_averages()` | Mn, Mw, Mz, dispersity |
| `step_sec_mw_distribution()` | d(W)/d(log M) curve |
| `step_sec_mw_fractions()` | Cumulative fractions at MW cutoffs |

### Stage 5: Export and Visualization

Extract results for reporting:

- `measure_sec_summary_table()`: Publication-ready summary statistics
- `measure_sec_slice_table()`: Point-by-point data for custom analysis
- `plot_sec_*()` family: Chromatograms, MWDs, calibration curves

## Choosing Your Calibration Approach

Calibration is where most users face their first major decision. This section helps you choose the right approach.

### Decision Tree

```{r decision-tree, echo = FALSE, fig.width=8, fig.height=6}
# Simplified decision tree visualization
library(tibble)

nodes <- tibble(
  x = c(4, 2, 6, 1, 3, 5, 7),
  y = c(4, 3, 3, 2, 2, 2, 2),
  label = c(
    "Do you have a\nlight scattering\ndetector?",
    "Do you have\nnarrow standards?",
    "Use MALS/LALS\nfor absolute MW",
    "Use broad\nstandard cal",
    "Same polymer\nas standards?",
    "",
    ""
  ),
  type = c("question", "question", "answer", "answer", "question", "answer", "answer")
)

edges <- tibble(
  x = c(4, 4, 2, 2, 3, 3),
  xend = c(2, 6, 1, 3, 3.8, 4.2),
  y = c(3.8, 3.8, 2.8, 2.8, 1.8, 1.8),
  yend = c(3.2, 3.2, 2.2, 2.2, 1.4, 1.4),
  label = c("No", "Yes", "No", "Yes", "No", "Yes")
)

# Add final answers
final_nodes <- tibble(
  x = c(3.8, 4.2),
  y = c(1.2, 1.2),
  label = c("Use universal\ncalibration", "Use conventional\ncalibration"),
  type = c("answer", "answer")
)

ggplot() +
  # Draw edges
  geom_segment(data = edges, aes(x = x, xend = xend, y = y, yend = yend),
               linewidth = 0.8, color = "gray50") +
  geom_text(data = edges, aes(x = (x + xend) / 2 + 0.15, y = (y + yend) / 2, label = label),
            size = 3, color = "gray30") +
  # Draw nodes

  geom_label(data = nodes |> dplyr::filter(type == "question"),
             aes(x = x, y = y, label = label),
             fill = "#E3F2FD", color = "#2E86AB", size = 3, label.padding = unit(0.4, "lines")) +
  geom_label(data = nodes |> dplyr::filter(type == "answer", label != ""),
             aes(x = x, y = y, label = label),
             fill = "#E8F5E9", color = "#1B5E20", size = 3, label.padding = unit(0.4, "lines")) +
  geom_label(data = final_nodes,
             aes(x = x, y = y, label = label),
             fill = "#E8F5E9", color = "#1B5E20", size = 3, label.padding = unit(0.4, "lines")) +
  labs(title = "Choosing Your Calibration Method") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold")) +
  coord_cartesian(xlim = c(0, 8), ylim = c(0.8, 4.5))
```

### Conventional Calibration

**When to use**: You have narrow MW standards (low dispersity) of the *same polymer type* as your samples.

**How it works**: Fit a polynomial relationship between log(MW) and elution time using standards. Apply this curve to unknown samples.

```r
step_sec_conventional_cal(
  standards_data = standards,
  degree = 3              # Polynomial order (usually 3)
)
```

**Advantages**: Simple, well-understood, widely used. Calibration curve provides diagnostics (R^2^, residuals).

**Limitations**: Assumes your sample behaves like the standards. A branched polymer will report incorrect MW because it elutes later than a linear polymer of the same MW.

**Best practices**:

- Use 5+ standards spanning your MW range of interest
- Check residuals---systematic patterns indicate poor fit
- Don't extrapolate far beyond your calibration range

### Broad Standard Calibration

**When to use**: You have one well-characterized broad-distribution reference material instead of multiple narrow standards.

**How it works**: Uses the known MWD of the reference to establish the MW-elution relationship in a single injection.

```r
step_sec_broad_standard(
  reference_mwd = known_distribution,
  reference_mn = 50000,
  reference_mw = 150000
)
```

**Advantages**: Single injection, good for routine QC. Convenient when narrow standards aren't available.

**Limitations**: Accuracy depends entirely on how well you know the reference material's MWD. Any error in the reference propagates to all samples.

### Universal Calibration

**When to use**: Your samples are a *different polymer type* than your available standards.

**How it works**: Calibrates on hydrodynamic volume ([η] × M) rather than molecular weight directly. Uses Mark-Houwink parameters to convert between polymer types.

```r
step_sec_universal_cal(
  # Calibration standards (e.g., polystyrene)
  reference_k = 1.14e-4,
  reference_a = 0.716,
  # Your sample (e.g., PMMA)
  sample_k = 6.0e-5,
  sample_a = 0.73
)
```

**Requires**: Intrinsic viscosity data (from viscometer) AND Mark-Houwink parameters for both polymer types.

**Advantages**: Accurate MW for different polymer types using common standards.

**Limitations**: You need reliable K and a values for your sample, which may require literature search or independent measurement. Doesn't work well for highly branched polymers where Mark-Houwink parameters aren't constant.

### Absolute MW (Light Scattering)

**When to use**: You need true molecular weight independent of polymer structure, or have no appropriate standards.

**How it works**: Light scattering intensity is directly proportional to molecular weight × concentration. No column calibration needed.

```r
step_sec_mals(
  dn_dc_column = "dn_dc"   # Refractive index increment
)
```

**Requires**: Light scattering detector (MALS, LALS, or RALS), accurate dn/dc value, good signal quality.

**Advantages**: True absolute MW. Works for any polymer type. Reveals branching through Rg-MW relationships. Required for proteins and complex polymers.

**Limitations**: More expensive instrumentation. Sensitive to dust and bubbles. Low MW samples may have weak scattering signal.

## Understanding Your Detectors

Each detector type provides different information. Knowing what each measures helps you interpret results and troubleshoot problems.

### Concentration Detectors

These tell you *how much* material is present at each elution slice.

#### Refractive Index (RI)

**What it measures**: Change in refractive index proportional to concentration.

**Key parameter**: dn/dc (refractive index increment, mL/g)

```r
step_sec_ri(dn_dc_column = "dn_dc")
```

| Pros | Cons |
|------|------|
| Universal---responds to all polymers | Sensitive to temperature |
| Mass-based response | Sensitive to mobile phase composition |
| Most common detector | Baseline drift common |

**When to use**: Default concentration detector for most polymer SEC. Essential for MW calculations.

#### UV Detector

**What it measures**: Absorbance at specific wavelength(s).

**Key parameter**: Extinction coefficient (ε, mL/(g·cm))

```r
step_sec_uv(extinction_column = "extinction_coef", wavelength = 280)
```

| Pros | Cons |
|------|------|
| Very sensitive | Only detects UV-active chromophores |
| Selective | Response depends on chromophore content |
| Stable baseline | May need multiple wavelengths |

**When to use**: Proteins (280 nm), aromatic polymers, or when combined with RI for composition analysis.

#### DAD (Diode Array)

**What it measures**: Full UV-Vis spectrum at each point.

```r
step_sec_dad(wavelengths = c(254, 280, 320))
```

**When to use**: Complex samples where spectral information helps identify components. Copolymer composition with wavelength-dependent absorption.

### Molecular Weight Detectors

These measure MW directly, independent of calibration.

#### Multi-Angle Light Scattering (MALS)

**What it measures**: Scattered light intensity at multiple angles.

**Provides**: Absolute MW and radius of gyration (Rg) at each slice.

```r
step_sec_mals(dn_dc_column = "dn_dc")
```

| Best for | Limitations |
|----------|-------------|
| Large molecules (> 10 kDa) | Requires very clean samples |
| Rg measurement | Signal weak for small molecules |
| Branching analysis | Expensive |

**See also**: [MALS Detection vignette](mals-detection.html)

#### Low-Angle Light Scattering (LALS)

**What it measures**: Scattered light near 0° angle.

```r
step_sec_lals(dn_dc_column = "dn_dc")
```

| Best for | Limitations |
|----------|-------------|
| Medium molecules | No Rg measurement |
| Simpler than MALS | Less accurate for large molecules |
| Lower cost | Still needs clean samples |

#### Right-Angle Light Scattering (RALS)

**What it measures**: Scattered light at 90°.

```r
step_sec_rals(dn_dc_column = "dn_dc")
```

| Best for | Limitations |
|----------|-------------|
| Small molecules | Biased for large molecules |
| Screening/QC | No structural information |
| Lowest cost LS option | Limited to narrow MW range |

**See also**: [LALS/RALS Detection vignette](lals-rals-detection.html)

### Hydrodynamic Detectors

These measure molecular size and conformation in solution.

#### Viscometer

**What it measures**: Differential pressure across a capillary, related to solution viscosity.

**Provides**: Intrinsic viscosity [η] when combined with concentration.

```r
step_sec_viscometer()
step_sec_intrinsic_visc(visc_col = "visc", conc_col = "concentration")
```

| Best for | Limitations |
|----------|-------------|
| Universal calibration | Requires concentration data |
| Branching characterization | Sensitive to bubbles |
| Mark-Houwink parameters | Additional complexity |

#### Dynamic Light Scattering (DLS)

**What it measures**: Time-dependent intensity fluctuations from Brownian motion.

**Provides**: Hydrodynamic radius (Rh).

```r
step_sec_dls()
```

| Best for | Limitations |
|----------|-------------|
| Particle sizing | Less common online |
| Conformation studies | Concentration-dependent |

## Common Pitfalls and Solutions

This section addresses problems you're likely to encounter and how `measure.sec` helps you fix them.

### Baseline Issues

**Symptoms**: Sloped baseline, wavy baseline, or baseline that doesn't return to zero.

**Causes**: Temperature drift (RI), mobile phase mixing, column equilibration, late-eluting contaminants.

**Solution**: Use `step_sec_baseline()` with an appropriate method:

```r
step_sec_baseline(
  measures = "ri",
  method = "linear",       # Simple start-to-end connection
  # or
  method = "spline",       # Follows gradual curves
  # or
  method = "asymmetric"    # Handles uneven noise
)
```

**Which method to choose**:

- **linear**: Fast, works when baseline is truly straight
- **spline**: Better for gradual drift or temperature effects
- **asymmetric**: Best for noisy data or when peak tails affect baseline

### Inter-Detector Delay

**Symptoms**: Calculated MW varies with injection volume, or MW appears to vary across a peak that should be uniform.

**Cause**: Detectors in series experience different delays. A 150 µL delay volume at 1 mL/min means 9 seconds of offset---which can shift MW by orders of magnitude.

**Solution**: Use `step_sec_detector_delay()` *before* any MW calculations:

```r
step_sec_detector_delay(
  reference = "ri",
  delay_volumes = c(uv = -0.05, mals = 0.15)  # mL
)
```

**How to measure delays**:

1. Inject a low-MW marker (toluene, polymer standard)
2. Record peak apex time on each detector
3. Convert time difference to volume: delay = Δt × flow_rate

**Tip**: Re-measure after column changes or major maintenance.

### Band Broadening

**Symptoms**: MWD appears broader than expected. High MW tail or low MW shoulder.

**Cause**: Peak spreading from extra-column volume, column dispersion, or injection effects.

**Solution**: Use `step_sec_band_broadening()` to deconvolve the true distribution:

```r
step_sec_band_broadening(
  sigma = 0.03  # Peak spreading parameter (mL)
)
```

**Estimate sigma** using a narrow standard---the peak width should be entirely from broadening:

```r
sigma <- estimate_sigma(
  narrow_standard_data,
  signal_col = "ri"
)
```

### Out-of-Range Calibration

**Symptoms**: Extrapolated MW values seem unreasonable. Very high or very low MW fractions.

**Cause**: Sample contains material outside the MW range of your calibration standards.

**Solution**: Options in `step_sec_conventional_cal()`:

```r
step_sec_conventional_cal(
  standards_data = standards,
  extrapolate = "linear",    # Continue slope beyond range
  # or
  extrapolate = "constant",  # Clamp at boundary
  # or
  extrapolate = "none"       # Return NA outside range
)
```

**Best practice**: Use standards that span your sample's MW range. If extrapolation is unavoidable, use "linear" with caution and flag those results.

### Noisy Light Scattering

**Symptoms**: Jagged LS signal, especially at low concentration regions.

**Causes**: Dust, bubbles, low sample concentration, detector noise.

**Solutions**:

1. **Filter samples**: 0.22 µm minimum, 0.1 µm for MALS
2. **Degas mobile phase**: Remove dissolved gases
3. **Increase concentration**: More sample = more scattering signal
4. **Apply smoothing**: In step parameters or post-processing

## Interpreting Results

### Molecular Weight Averages

SEC produces multiple MW averages. Each tells you something different:

| Average | Formula | What It Emphasizes |
|---------|---------|-------------------|
| Mn | Σ(n~i~M~i~) / Σn~i~ | Number of chains---sensitive to low MW |
| Mw | Σ(n~i~M~i~^2^) / Σ(n~i~M~i~) | Mass of chains---sensitive to high MW |
| Mz | Σ(n~i~M~i~^3^) / Σ(n~i~M~i~^2^) | Very sensitive to high MW tail |

**Practical interpretation**:

- **Mn** tells you about the most numerous species
- **Mw** tells you where most of the mass is
- **Mz** warns about high MW contamination or aggregation

### Dispersity (Đ)

Dispersity = Mw/Mn measures the breadth of the distribution.

| Đ Value | Interpretation |
|---------|----------------|
| 1.0 | Perfectly monodisperse (theoretical limit) |
| < 1.1 | Very narrow (controlled polymerization) |
| 1.5 - 2.0 | Typical for step-growth or conventional free radical |
| > 3.0 | Very broad, may indicate blending or degradation |

**Tip**: If Đ < 1.0, something is wrong---often a calibration or integration problem.

### When Results Look Wrong

**MW seems too high or too low**:

- Check calibration: Are standards correctly identified?
- Check detector delays: Misalignment shifts apparent MW
- Check dn/dc: Wrong value directly scales MW from LS

**Dispersity seems too narrow**:

- Check integration limits: Too narrow excludes real distribution
- Check band broadening: Unaccounted broadening inflates apparent width
- Sample may genuinely be narrow

**MW varies between injections**:

- Check baseline reproducibility
- Check column equilibration
- Check sample stability (aggregation, degradation)

**Negative MW at distribution edges**:

- Normal at very low signal---noise can go negative
- Set appropriate thresholds in MW calculation steps
- Focus on the region with good signal-to-noise

## See Also

- [Getting Started](getting-started.html) - Your first SEC analysis
- [Comprehensive SEC Analysis](sec-analysis.html) - Full reference for all features
- [Multi-Detector SEC](triple-detection.html) - Detailed multi-detector workflows
- [Calibration Management](calibration-management.html) - Saving and reusing calibrations
